Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-31 introduction.tex
proof

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

Various forms of \emph{reproduction number} were prominent throughout the SARS-CoV-2 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation}, and building on over a century of infectious disease dynamics literature, I define three that are broadly useful.

@@

\emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
It is the primary metric for the magnitude of a pandemic .
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths); although this requires a severity measure, such as the IFR (infection fatality rate), the proportion of infected individuals who die.

@@

Data on severe disease is typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

However, severe disease can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
Therefore, the data does not provide much information on incidence in those subgroups.
Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.

@@

Prevalence can be estimated through a \emph{prevalence survey}.
A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
On the other hand, they are a source of prevalence-type data that can be unbiased.
For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.

@@

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large cohort to be appropriately powered.
Large cohort studies are expensive to run.

@@

The models with the weakest assumptions are \emph{phenomenological} models.
These make almost no assumptions about the transmission process, viewing incidence as resulting from a counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
A common model is the Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
The model parameters can be used to define an effective reproduction number as either $\Re = \int_0^\infty \beta(t, \tau) d\tau$ or $\Re = \int_0^\infty \beta(t+\tau, \tau) d\tau$~\autocite{gosticPractical}.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.

@@

The infection process is hardly ever directly observed; however, we often observe a second process.
\emph{Disease models} link the transmission model to the value, $W_t$ of this second process at time $t$.
For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section), which could introduce additional complications.

@@

Observation models link $W_t$ to a possibly biased and/or noisy observation of it at time $t$.

@@

Modelling the duration directly considers only the binary result at each test (positive or negative).
This approach generally requires fewer parameters reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is useful in situation where the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
More details are in \cref{E-perf-test}.

@@

The surveillance systems used in response to SARS-CoV-2 are unique. 
The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing; severe events; seroprevalence and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.

@@

Severe events counts were generally a more reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded.
Therefore, these data were subject to fewer biases than the cases detected through community testing.
However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and attempt to observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

@@

A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
This meant that using REACT data to infer transmission in these periods requires strong model assumptions.
A particular instance of this is that a peak in prevalence in late December 2020 occurred between rounds of the REACT survey, complcating any analysis of REACT data over that period~\autocite{rileyREACTround8}.

@@

One of the studies, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
ATACCC recruited individuals from contacts of known infections reported to NHS Test and Trace, and collected data on them for 20 days.
This included daily RT-PCR tests, which allowed quantification of their viral load (the amount of virus in an individual), a biomarker.
The density of the sampling allows for modelling the trajectory of viral load to estimate the duration of detectability (how long an individual is RT-PCR positive for).
However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.

@@

The massive impact of the SARS-CoV-2 pandemic led to data collection on a scale never previously seen for an infectious disease.
Analysis of this data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
Therefore, a deep understanding of the potential of each data source was not possible.
In this thesis I investigate the potential of the CIS as a single source of information for incidence and transmission estimation.

@@

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
It specifies a particular combination of infection, disease, and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
The theory behind the framework underpins much of the work in this thesis.

The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a Poisson process as the infection model; a disease model with a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis and assumed $\zeta=$; and a trivial observation model where all infections are (eventually) detected.

In (see \cref{E-inc-prev}) I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
The use of prevalence-type data induces a dependency structure between the values of the latent process $W_1, \dots, W_t$; I include the dependency by modifying the disease model.
In addition, unlike in the original formulations, not all infections are detected by the CIS.
In the CIS, a sample of the population receive RT-PCR tests, not the entire population; I include the sampling error arising from this in the observation model.

@@

In the application of the backcalculation to the CIS data, the delay function $f$ represents the duration of detectability.
Previous work has attempted to estimate $f$.
An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The longest reported observed duration was 83 days, although no individuals were infectious beyond day nine.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.

@@

\Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
Taking the time of the initial positive test as time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design does not provide any information on how long before the initial positive test the infection occurred.
This issue also means the study underestimates the duration of detectability, because the infection's true start time would be prior to the first positive test.

For the purposes of this thesis, I require estimates of the duration of detectability applicable to the general population.
As no such estimates of the full distribution of the duration are available in the literature, a key component of this work is devoted to estimating them.
To do so, I will combine the information from the ATACCC and CIS studies.

\subsection{Existing estimation from prevalence surveys}

@@

To address this question, I analyse data from the CIS study.
I use, as a case study, data from the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
This period has several interesting features making it suitable for my aim.
First, the CIS was of sufficient size and transmission was sufficiently high to provide enough data to analyse.
Second, incidence began to rise over this period, after the fall due to the first lockdown.
Finally, the period ends before the vaccine roll-out, which  affected many important aspects of the analysis; most relevantly, the duration of detectability and the susceptibility of the population.
Incorporating vaccination effects would require more complex modelling and include idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.

@@

    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence estimation.
    Furthermore, I use my estimate of the duration of detectability from general population studies.

    \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang---Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
    I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
    I show the CIS data are sufficient for estimating incidence and transmission.
\end{enumerate}

@@

Almost all code, excluding dependencies, was written from scratch for these analyses.
The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevalence used in \cref{E-backcalc}, and the original viral load model used in \cref{E-ATACCC:sec:hakki}.

@@

In \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
I show the distribution of the duration of detectability is the key quantity linking incidence and prevalence.

In \cref{E-ATACCC}, I estimate the duration of detectability from the ATACCC study.
This provides a reliable estimate of the bulk of the distribution (out to around 20 days),  but shows that estimating the tail requires a large sample size and long follow-up. 
These features are present in the CIS, in \cref{E-perf-test} I provide a framework for estimating the duration of detectability assuming perfect tests from prevalence surveys such as the CIS.
However, as the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.

@@

Finally, I conclude in \cref{E-conclusion} with a discussion of the limitations of this work and ideas for future research.

>>>>

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have, on average, occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

Various forms of \emph{reproduction number} were prominent throughout the SARS-CoV-2 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation}, and building on over a century of infectious disease dynamics literature, I define three below.

@@

\emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
It is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths) through a severity measure, such as the infection fatality rate, the proportion of infected individuals who die.

@@

Data on the severe consequences of an infection typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

However, severe consequences can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
Therefore, the data does not provide much information on incidence in those subgroups.
Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.

@@

Prevalence can be estimated through a \emph{prevalence survey}.
A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately but can provide unbiased prevalence-type data.
An example of an unbiased prevalence survey is testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.

@@

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large sample size to be appropriately powered, making them expensive to run.

@@

The models with the weakest assumptions are \emph{phenomenological} models.
These make almost no assumptions about the transmission process, viewing incidence as resulting from a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}), such as a Poisson process.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal}.
In discrete time, it takes the form:
\begin{align}
    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=0}^t \beta(t,\tau) Z_{t-\tau} d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z_t$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
The model parameters can be used to define an effective reproduction number as either $\Re = \sum{\tau=0}^\infty \beta(t, \tau)$ or $\Re = \sum_{\tau=0}^\infty \beta(t+\tau, \tau)$~\autocite{gosticPractical}.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.

@@

The infection process is hardly ever directly observed; however, we often observe a consequence of an infection.
\emph{Disease models} link the infection model to the number of these consequences, $W_t$ at time $t$.
For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see the next section).

@@

Observation models link $W_t$ to a (possibly biased and/or noisy) observation of it at time $t$.

@@

Modelling the duration directly considers only the binary result at each test (positive or negative), using less information than a biomarker approach.
This approach generally requires fewer parameters reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is useful in situation where the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
More details are in \cref{E-perf-test}.

@@

The surveillance systems set-up in response to SARS-CoV-2 were on a scale never previously seen.
The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing, severe events, seroprevalence surveys, and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.

@@

The number of severe events was generally a more reliable measure of the burden of the disease, subject to fewer changes than the number of cases detected through community testing.
However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

@@

A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
This meant that using REACT data to infer transmission in these periods requires strong model assumptions.
A particular instance of this is that a peak in prevalence in late December 2020 occurred between rounds of the REACT survey, complicating any analysis of REACT data over that period~\autocite{rileyREACTround8}.

@@

Another study, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
ATACCC recruited individuals from contacts of known infections reported to NHS Test and Trace, and collected data on them for 20 days.
This included daily RT-PCR tests, which allowed quantification of their viral load (the amount of virus in an individual), a biomarker.
The density of the sampling allows for modelling the trajectory of viral load to estimate the duration of detectability (how long an individual is RT-PCR positive for).
However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.

@@

Analysis of pandemic data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
Therefore, a deep understanding of the potential of each data source was not possible.
In this thesis I investigate the potential of the CIS as a single source of information for incidence and transmission estimation.

@@

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
It specifies a particular combination of infection, disease, and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
I build on the theory behind the framework in this thesis.

The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a phenomenological Poisson process as the infection model; a disease model with a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis and assumed $\zeta=1$; and a trivial observation model where all diagnoses are recorded.

In \cref{E-inc-prev}, I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
The use of prevalence-type data induces a dependency structure between the values of the latent process $W_1, \dots, W_t$; I include the dependency by modifying the disease model.
In addition, unlike in the original formulations, not all infections are detected by the CIS.
In the CIS, a sample of the population receive RT-PCR tests, not the entire population; I include the sampling error arising from this in the observation model.

@@

In the application of the backcalculation to the CIS data, the delay function $f$ represents the duration of detectability.
Previous work has attempted to estimate $f$.

An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The longest reported observed duration was 83 days, although no individuals were infectious beyond day 9.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
In addition, summary estimates are insufficient for backcalculation, which requires a full distribution.

@@

\Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6--22 Jan 2021) were invited to take two additional swab tests.
Taking the time of the initial positive test as time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
Strong assumptions are necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prevents assessing this hypothesis because the design does not provide any information on how long before the initial positive test the infection occurred.
This issue also means the study underestimates the duration of detectability, because the infection's true start time would be prior to the initial positive test.

For the purposes of this thesis, I require estimates of the duration of detectability applicable to the general population.
As no such estimates of the full distribution of the duration are available in the literature, a key component of this thesis is estimating them.
To do so, I will combine the information from the ATACCC and CIS studies.

\subsection{Existing use of prevalence surveys}

@@

To address this question, I use a case study.
Specifically, the CIS data from the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
This case study is suitable for addressing this question for several reasons.
First, the CIS was of sufficient size, and transmission was sufficiently high, to provide enough data for the analyses.
Second, incidence began to rise over this period, after the fall due to the first lockdown.
Finally, the period ends before the vaccine roll-out, which  affected many important aspects of the analysis; most relevantly, the duration of detectability and the susceptibility of the population.
Incorporating vaccination effects would require more complex modelling and include idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.

@@

    \item In \cref{E-backcalc}, I use a deterministic, phenomenological backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence survey.
    Furthermore, I use my estimate of the duration of detectability, which is from general population studies.

    \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang--Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
    I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
    I show the CIS data are sufficient for estimating incidence and transmission.
\end{enumerate}

@@

Almost all code, excluding dependencies, was written from scratch for these analyses.
The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevalence used in \cref{E-biology-data:sec:MRP}, and the original viral load model used in \cref{E-ATACCC:sec:hakki}.

@@

In \cref{E-inc-prev}, I formally describe the infection, disease, and observation models for a prevalence survey.
I show the distribution of the duration of detectability is the key quantity linking incidence and prevalence.

In \cref{E-ATACCC}, I estimate the duration of detectability from the ATACCC study.
This provides a reliable estimate of the bulk of the distribution (out to around 20 days),  but shows that estimating the tail requires a large sample size and long follow-up. 
These features are present in the CIS.
In \cref{E-perf-test}, I provide a framework for estimating the duration of detectability assuming perfect tests from prevalence surveys such as the CIS.
However, as the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.

@@

Finally, I conclude in \cref{E-conclusion} with a discussion of my findings, the limitations of this work, and ideas for future research.

### 2024-03-31 thesis.tex
proof

\abstract{%
    Surveillance is a crucial component of a pandemic response, informing public health and wider policy.
    Unfortunately, it is often subject to reporting and behavioural biases, reducing its robustness.
    % During the SARS-CoV-2 epidemic, the UK government funded two large-scale prevalence surveys with samples recruited randomly from the general population.
    % One of these, the CIS (Coronavirus Infection Survey), was a longitudinal survey run from April 2020 to March 2023.
    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey, with participants recruited randomly from the general population, thereby avoiding these issues.

@@

    This relies on knowledge of the duration of detectability (how long individuals test positive for).
    Therefore, I first consider how to estimate the distribution of this duration in the general population.
    I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of detectability.
    % I take two complementary approaches using different datasets.
    % This model is fitted to data from a longitudinal study of contacts of individuals with confirmed SARS-CoV-2 infection.
    % In \cref{ATACCC}, I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of detectability.
    % This model is fitted to data from the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study, a longitudinal study of viral load in contacts of individuals with confirmed SARS-CoV-2 infection.
    I use this detailed model to exploit frequent (daily) sampling of individuals in a cohort of contact traced individuals.
    However, the cohort was small and the follow-up time short meaning that the tail of the duration distribution is poorly estimated.

    % Therefore, in \cref{perf-test,imperf-test}, I combine the information from ATACCC with the larger CIS dataset.
    % I analyse the CIS in a survival analysis framework.
    To overcome this issue, I combine the information with the larger and longer CIS dataset.
    The CIS data poses several challenges because it is doubly interval censored and some individuals in the cohort have infections that are never detected.
    I apply a survival analysis framework to analyse this data, which had previously only been used in simpler settings.
    In addition, I develop novel methodology to incorporate false negatives into the framework.
    % The results showed that false negatives in the CIS data needed to be included in the model.
    % In \cref{imperf-test} I develop a novel methodology to do so.
    This results in an estimate of the full duration distribution in the general population, not previously available.

    Finally, I use this estimate to infer the incidence and transmission of SARS-CoV-2 in England from the CIS.
    I fully propagate the sampling error, accurately include the shape of the duration distribution, and estimate the quantities by age; previous work lacked at least one of these.
    Furthermore, I fit to only the CIS data, showing that such a survey is adequate to estimate these quantities.
    % without requiring an evidence synthesis approach previously used.
}

@@

    Finally, I must thank my family, especially my parents, Laurence and Tracy Blake.
    Throughout my life, no ask has ever been too much.
    Your love and support are infinite and unconditional, your gentle nudges towards the right path irreplaceable.
    Thank you for everything.

    Having written this, I realise how fortunate I am to have been supported by such great people.
    I can never thank you enough.
}

@@

    Several aspects of the work in this thesis were completed in support of the ongoing response.
    In particular, an earlier version of the phenomenological backcalculation was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
    This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.

@@

    During this period, I have been involved in several other pieces of work and discussion contributing to the response, not directly used in this thesis.
    These have led to several publications, listed in \cref{publications} for completeness.

>>>>

\abstract{%
    Surveillance is a crucial component of a pandemic response, informing public health and wider policy.
    Unfortunately, it is often subject to reporting and behavioural biases, reducing its robustness.
    % During the SARS-CoV-2 epidemic, the UK government funded two large-scale prevalence surveys with samples recruited randomly from the general population.
    % One of these, the CIS (Coronavirus Infection Survey), was a longitudinal survey run from April 2020 to March 2023.
    The CIS (Coronavirus Infection Survey) was a large-scale, longitudinal prevalence survey, with participants recruited randomly from the general population, thereby avoiding these issues.

@@

    This relies on knowledge of the duration of detectability (how long individuals test positive for).
    Therefore, I first consider how to estimate the distribution of this duration in the general population.
    I adapt a pre-existing model of how an individual's viral load changes over time for this purpose.
    % I take two complementary approaches using different datasets.
    % This model is fitted to data from a longitudinal study of contacts of individuals with confirmed SARS-CoV-2 infection.
    % In \cref{ATACCC}, I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of detectability.
    % This model is fitted to data from the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study, a longitudinal study of viral load in contacts of individuals with confirmed SARS-CoV-2 infection.
    I use this detailed model to exploit frequent (daily) sampling of individuals in a cohort of contact traced individuals.
    The cohort was small and the follow-up time short meaning that the tail of the duration distribution is poorly estimated.

    % Therefore, in \cref{perf-test,imperf-test}, I combine the information from ATACCC with the larger CIS dataset.
    % I analyse the CIS in a survival analysis framework.
    To overcome this issue, I combine the information with the larger and longer CIS dataset.
    The CIS data poses several challenges because it is doubly interval censored and some individuals in the cohort have infections that are never detected.
    I develop a survival analysis framework to analyse this data, including novel methodology to incorporate false negatives into the framework.
    % The results showed that false negatives in the CIS data needed to be included in the model.
    % In \cref{imperf-test} I develop a novel methodology to do so.
    This results in an estimate of the full duration distribution in the general population, not previously available.

    Finally, I use this estimate to infer the incidence and transmission of SARS-CoV-2 in England, using only the CIS data.
    I fully propagate the sampling error, accurately include the shape of the duration distribution, and estimate the quantities by age; previous work lacked at least one of these.
    Furthermore, I fit to only the CIS data, showing that such a survey is adequate to estimate these quantities.
    % without requiring an evidence synthesis approach previously used.
}

@@

    Finally, I must thank my family, especially my parents, Laurence and Tracy Blake.
    Throughout my life, no ask has ever been too much.
    Your love and support are infinite and unconditional, your gentle nudges towards the right path irreplaceable.
    Thank you for everything.
}

@@

    Several aspects of the work in this thesis were completed in support of the ongoing response.
    In particular, an earlier version of the phenomenological backcalculation developed in \cref{backcalc} was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
    This method was also used for an analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.

@@

    During this period, I have been involved in several other pieces of work and discussion in response to the pandemic, not directly used in this thesis.
    These have led to several publications, listed in \cref{publications} for completeness.
